---
title: "p8105_hw2_xx2554"
author: "Xinyan Xu"
output: github_document
date: "2026-10-05"
---

```{r setup, message = FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r}
transit_df = 
  read_csv(
    "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character()
    )
  ) |> 
  janitor::clean_names() |> 
  select(
    line, station_name, station_latitude, station_longitude,
    starts_with("route"), entry, vending, entrance_type, ada
  ) |> 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    ))

transit_df
```


This dataset contains information on each entrance and exit of the NYC subway system, so each row represents one entrance or exit rather than one station. The cleaned dataset includes line, station name, station latitude and longitude, the routes served (`route1` to `route11`), entry, vending, entrance type, and ADA compliance. To clean the data, I imported the CSV using a relative path (reading `route8` to `route11` as character so all route columns have the same type), cleaned variable names with `janitor::clean_names()`, kept the variables of interest with `select`, and converted `entry` from `YES`/`NO` to a logical variable using `case_match`. The resulting dataset has `r nrow(transit_df)` rows and `r ncol(transit_df)` columns. These data are not tidy: route served is a single variable, but it is spread across 11 columns with the route number stored in the column names, and most of these columns are missing because most stations serve only a few routes.

### Distinct stations

```{r}
n_stations = 
  transit_df |> 
  distinct(line, station_name) |> 
  nrow()
```

There are `r n_stations` distinct stations, identified by both station name and line.

### ADA compliant stations

```{r}
n_ada_stations = 
  transit_df |> 
  filter(ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
```

`r n_ada_stations` stations are ADA compliant.

### Entrances without vending that allow entrance

```{r}
prop_entry_no_vending = 
  transit_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

Among station entrances / exits without vending (counted at the entrance level, not the station level), `r round(prop_entry_no_vending * 100, 1)`% allow entrance.

### Reformatting routes

```{r}
transit_tidy_df = 
  transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |> 
  drop_na(route_name) |> 
  mutate(route_number = as.numeric(route_number))

transit_tidy_df
```

The reformatted dataset has one row for each route served at each entrance / exit, with `route_number` and `route_name` as separate variables (`r nrow(transit_tidy_df)` rows and `r ncol(transit_tidy_df)` columns).

### Stations serving the A train

```{r}
n_a_stations = 
  transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

n_a_ada_stations = 
  transit_tidy_df |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(line, station_name) |> 
  nrow()
```

`r n_a_stations` distinct stations serve the A train, and `r n_a_ada_stations` of these are ADA compliant.


## Problem 2

### Mr. Trash Wheel

```{r}
mr_trash_df = 
  read_excel(
    "./data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N743") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "mr_trash_wheel")

mr_trash_df
```

### Professor Trash Wheel and Gwynnda

```{r}
professor_trash_df = 
  read_excel(
    "./data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M140",
    na = c("", "NA", "dive")) |>  
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(trash_wheel = "professor_trash_wheel")

gwynnda_trash_df = 
  read_excel(
    "./data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(trash_wheel = "gwynnda")

professor_trash_df
gwynnda_trash_df
```


### Combine datasets

```{r}
trash_wheel_df = 
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_trash_df) |> 
  mutate(month = str_to_title(str_trim(month))) |> 
  select(trash_wheel, everything())

trash_wheel_df
```

### Description

```{r}
prof_total_weight = 
  trash_wheel_df |> 
  filter(trash_wheel == "professor_trash_wheel") |> 
  pull(weight_tons) |> 
  sum(na.rm = TRUE)

gwynnda_june_2022_butts = 
  trash_wheel_df |> 
  filter(trash_wheel == "gwynnda", year == 2022, month == "June") |> 
  pull(cigarette_butts) |> 
  sum(na.rm = TRUE)
```

The combined dataset contains `r nrow(trash_wheel_df)` observations, where each row represents one dumpster collected by Mr. Trash Wheel (`r nrow(mr_trash_df)` dumpsters), Professor Trash Wheel (`r nrow(professor_trash_df)` dumpsters), or Gwynnda (`r nrow(gwynnda_trash_df)` dumpsters) between `r min(trash_wheel_df$year)` and `r max(trash_wheel_df$year)`. Key variables include `trash_wheel`, the collection `date`, `weight_tons`, `volume_cubic_yards`, and counts of trash types such as `plastic_bottles` and `cigarette_butts`; some trash types (e.g. `sports_balls`) are only recorded for certain trash wheels, so these are missing for the others. For available data, Professor Trash Wheel collected a total of `r prof_total_weight` tons of trash, and Gwynnda collected `r format(gwynnda_june_2022_butts, big.mark = ",")` cigarette butts in June of 2022.


## Problem 3

### Baseline demographics

```{r}
baseline_all_df = 
  read_csv(
    "./data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"),
    sex = as.factor(sex),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non_carrier"),
    apoe4 = as.factor(apoe4))

baseline_df = 
  baseline_all_df |> 
  filter(is.na(age_at_onset) | age_at_onset > current_age)

baseline_df
```

### Baseline description and questions

```{r}
n_recruited = nrow(baseline_all_df)

n_eligible = nrow(baseline_df)

n_develop_mci = 
  baseline_df |> 
  drop_na(age_at_onset) |> 
  nrow()

mean_baseline_age = 
  baseline_df |> 
  pull(current_age) |> 
  mean()

prop_women_apoe4 = 
  baseline_df |> 
  filter(sex == "female") |> 
  mutate(is_carrier = apoe4 == "carrier") |> 
  pull(is_carrier) |> 
  mean()
```

The baseline data were imported with `skip = 1` (the first row contains variable descriptions) and with `.` treated as missing, since age at onset is recorded as `.` for participants who remained MCI free. `sex` and `apoe4` were converted from 1/0 codes to labeled factors. Each row is one participant, with variables including baseline age, sex, education, APOE4 status, and age at MCI onset. A total of `r n_recruited` participants were recruited; `r n_recruited - n_eligible` were removed because their age at onset was at or before their baseline age (i.e. they already had MCI), leaving `r n_eligible` participants, of whom `r n_develop_mci` developed MCI during follow-up. The average baseline age is `r round(mean_baseline_age, 2)` years, and `r round(prop_women_apoe4 * 100, 1)`% of women are APOE4 carriers.



